A comparison of single-cell trajectory inference methods: towards more accurate and robust tools

### 2018-09-17 manuscript/paper.Rmd
update paper for v3, it builds! :confetti:

---
title: "A comparison of single-cell trajectory inference methods: towards more accurate and robust tools"
bibliography: references.bib
csl: nature-biotechnology.csl
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage[table]{xcolor}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{tabu}
  - \usepackage[normalem]{ulem}
  - \usepackage{lscape}
  - \usepackage{pbox}
  - \usepackage[labelformat=empty]{caption}
  - \usepackage{hyperref}
  - \definecolor{airforceblue}{rgb}{0.36, 0.54, 0.66}
  - \definecolor{cerulean}{rgb}{0.0, 0.48, 0.65}
  - \definecolor{burntorange}{rgb}{0.8, 0.33, 0.0}
  - \hypersetup{colorlinks=true,linkcolor=airforceblue,citecolor=cerulean,urlcolor=burntorange}
  - \ExecuteBibliographyOptions{sorting=none,sortcites=true,autocite=superscript,autopunct=false}
  - \newcommand{\hideFromPandoc}[1]{#1}
  - \hideFromPandoc{\let\Begin\begin}
  - \hideFromPandoc{\let\End\end}
  - \hideFromPandoc{\newenvironment{myfigure}[1]{\begin{figure}[#1]}{\end{figure}}}
mainfont: Open Sans

---

```{r setup, include=FALSE}
source(paste0(dynalysis::get_dynalysis_folder(), "/analysis/paper/setup.R"))

# knitr options
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = paste0(rprojroot::find_rstudio_root_file(), "/analysis/paper/.scratch/")
)
```

@@

# Abstract

Recent technological advances allow unbiased investigation of cellular dynamic processes, by generating -omics profiles of thousands of single cells and computationally ordering these cells along a trajectory. Since 2014, at least `r sum(implementations$contains_ti, na.rm=TRUE)` trajectory inference methods have been developed, but due to high variability in their inputs and outputs, these methods are difficult to compare. As a result, a comprehensive assessment of the performance of trajectory inference methods is still lacking, and more importantly also guidelines for new users to the field.

In this study, we evaluated a total of  `r nrow(filter(methods, evaluated, type == "algorithm"))` methods in terms of cellular ordering, trajectory topology, code quality and user friendliness. Our results indicate that existing methods are typically better at ordering the cells rather than inferring the correct trajectory topology, and that there is no "one-fits-all" method. Instead, the choice of method depends on the trajectory topology present in the data. We therefore developed a set of guidelines to aid users select the most appropriate method, based on their prior knowledge of the trajectory present in the data. While the field certainly has matured significantly since its inception, further progress is still necessary in order to cope with increasingly complex and novel use cases. We hope to spearhead such developments and the field as a whole, by making our evaluation pipeline easily extensible and freely available at [github.com/dynverse/dynverse](https://www.github.com/dynverse/dynverse).

@@

<!-- §2: Plethora of new TI methods -->
A plethora of TI methods has been developed over the last years, and even more are being created every month (`r ref("stable", "implementations")`). For no other single-cell analysis workflows have so many tools been developed, according to several repositories such as omictools.org [@henry_omictoolsinformativedirectory_2014], the "awesome single cell software" list [@davis_awesomesinglecelllistsoftware_2018] and scRNA-tools.org [@zappia_exploringsinglecellrnaseq_2017]. All methods have their own unique set of characteristics in terms of underlying algorithm, the prior information they require and the outputs they produce. Two of the most distinctive differences between TI methods are whether they fix the topology of the trajectory or not, and what type of topologies they can detect. Early TI methods typically fixed the topology algorithmically (e.g. linear [@bendall_singlecelltrajectorydetection_2014; @schlitzer_identificationcdc1cdc2committed_2015; @shin_singlecellrnaseqwaterfall_2015; @campbell_bayesiangaussianprocess_2015] or bifurcating [@haghverdi_diffusionpseudotimerobustly_2016; @setty_wishboneidentifiesbifurcating_2016]), or through parameters provided by the user [@trapnell_dynamicsregulatorscell_2014; @matsumoto_scoupprobabilisticmodel_2016]. These methods therefore mainly focused on correctly ordering the cells along the fixed topology. 
Other methods attempt to infer the topology computationally, which increases the difficulty of the problem at hand, but allows these methods to be applicable on more use cases. Methods that perform topology inference are still in the minority, though current trends suggest this will change soon.

```{r, results='asis' }
add_stable(
  read_rds(figure_file("implementations_table.rds", "4-method_characterisation"))[[params$table_format]],
"implementations",
paste0("Overview of the trajectory inference methods included in this study. ", glue::glue_data(non_inclusion_reasons, "{footnote}: {long}") %>% glue::collapse("; "))
)
```

@@

<!-- §4: Our study -->
In this study, we performed a comprehensive evaluation (`r ref("fig", "ti_evaluation_overview", "a")`, `r ref("sfig", "overview_evaluation_expanded")`) of  `r nrow(filter(methods, evaluated, type == "algorithm"))` TI methods. The inclusion criterion for TI methods was primarily based on their free availability and presence of a programming interface (`r ref("stable", "implementations")`). In addition, we added three control methods to define baseline performance: a random trajectory, Component 1 as a control for linear trajectories and Angle as a control for circular trajectories. The evaluation comprised two core aspects: (i) assessment of the accuracy and scalability of TI methods by comparing predicted trajectories with a gold standard on `r sum(tasks_info$task_source == "real")` real and `r sum(tasks_info$task_source == "synthetic")` synthetic datasets, and (ii) a quality control of the provided software and documentation. Finally, we combined both the performance and QC evaluation to provide a set of practical guidelines for users.

```{r echo=FALSE, results="asis"}
add_fig(
  fig_path = figure_file("overview_evaluation_v4.svg", experiment_id="manual_figures"),
  ref_id = "ti_evaluation_overview", 
  caption_main = "Overview of several key aspects of the evaluation.", 
  caption_text = "a) A schematic overview of our evaluation pipeline. b) In order to make the trajectories comparable to each other, a common trajectory model was used to represent gold standard trajectories from the real and synthetic datasets, as well as any predicted trajectories from TI methods. c) Trajectories are automatically classified into one of nine trajectory types. d) Three metrics were defined in order to assess the similarity in cell ordering, cell neighbourhood, and topology, for any two trajectories."
)
```

```{r}
add_sfig(
  fig_path = figure_file("overview_evaluation_expanded_v1.svg", "manual_figures"),
  ref_id = "overview_evaluation_expanded",
  caption_main = "Extended overview of our evaluation pipeline.", 
  caption_text = "From literature we extracted a quality control checklist, a list of trajectory inference methods, a set of real datasets containing a trajectory and real regulatory networks used to generate the synthetic data. We created a wrapper of each method, so that its output was transformed into a common probabilistic trajectory model, and used these to infer trajectories on all real and synthetic datasets using their default parameters. Using several similarity metrics, we compared the gold standard of the real and synthetic datasets with the inferred trajectories. We also evaluated the quality of each method using the quality control checklist, and used both the benchmark results and quality control results to produce a final set of guidelines for method's users.", 
  width = 15, 
  height = 7.5
)
```

@@

<!-- Evaluation overview -->
In order to make gold standard trajectories and predicted trajectories directly comparable to each other, we developed a common probabilistic model for representing trajectories from all possible sources (`r ref("fig", "ti_evaluation_overview", "b")`). In this model, the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of connected milestones. The milestone network is automatically classified into one of the nine predefined trajectory types (`r ref("fig", "ti_evaluation_overview", "c")`), which go from the very basic topologies (linear, cyclical and bifurcating) to the most complex (cyclic graphs and disconnected graphs). We defined a set of metrics for calculating the similarity of two trajectories represented in the common model, each assessing a different aspect of a trajectory: the cell ordering, the cell neighbourhood and the topology (`r ref("fig", "ti_evaluation_overview", "d")`). The data compendium consisted of both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance. These real datasets come from a variety of single-cell technologies, organisms, and dynamic processes, and contain several types of trajectory topologies (`r ref("stable", "datasets")`). In order to generate synthetic datasets, we simulated gene regulatory networks using a thermodynamic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and subsequently simulated single-cell profiling experiments by matching the distributions of the synthetic data with real reference datasets.

<!-- Overall evaluation results -->
Across all datasets and evaluation metrics, we found that Slingshot, TSCAN, SCORPIUS, AGA pseudotime and cellTree predicted the most accurate trajectories (`r ref("fig", "results_overview", "b")`).  When focusing on the benchmark scores per trajectory type, we found that several methods were specialised in predicting specific trajectory types; for example SCORPIUS and cellTree maptpx returned the most accurate linear trajectories, Angle and reCAT optimally identified cycles, and Monocle DDRTree, SCOUP and AGA pseudotime most accurately identified trees. This was also reflected in the rankings of the different metrics, with methods able to detect more complex trajectories, such as DPT, TSCAN, SCOUP and AGA, obtaining a better topological similarity score, while other methods, including SCORPIUS, Slingshot, Angle and cellTree, were better at predicting the ordering and neighbourhood of the cells (`r ref("fig", "results_overview", "b")`).

```{r echo=FALSE, results="asis"}
add_fig(
  fig_path = figure_file("overview.svg", experiment_id = "9-main_figure"),
  ref_id = "results_overview", 
  caption_main = glue::glue("Overview of the results on the evaluation of {nrow(filter(methods, evaluated, type == 'algorithm'))} TI methods and {nrow(filter(methods, evaluated, type == 'control'))} control methods."), 
  caption_text = "a) The methods were characterised according to the most complex trajectory type they can infer, whether or not the inferred topology is constrained by the algorithm or a parameter, and which prior information a method requires or can optionally use. b) The methods are ordered according to the overall score achieved both in the benchmark and QC evaluation. Three controls (random trajectory, component 1 of a PCA and angle on a PCA) were added, which provide a baseline performance. c) Also shown are the aggregated scores per metric, source and trajectory type, as well as the average execution time across all datasets and the percentage of executions in which no output was produced. d) Performance in the quality control evaluation is highly variable, even amongst the highest ranked methods according to the benchmark evaluation. Also listed are the quality control scores aggregated according to practicality and the different categories."
)
```

<!-- Errors-->
Several methods obtained a high overall score despite having 5 to 10% of failed executions (e.g. Monocle DDRTree, cellTree Gibbs, SLICE), meaning these methods could rank even higher if not for the failed executions. Other methods (reCAT, SCOUP, StemID, cellTree Gibbs) exceeded the pre-defined time or memory limits when applied to the largest datasets, and therefore were not scalable to large numbers of cells or genes. For a few methods, the time or memory limits were exceeded too often, making the benchmarking results difficult to compare to those of other methods (Ouija, Pseudogp).

<!-- Evaluation robustness -->
```{r}

# the correlation when subsampling
correlation_tenth <- read_rds(result_file("task_sample_correlations.rds", "11-evaluation_robustness")) %>%
filter(n_tasks == round(max(n_tasks)/10)) %>% 
          pull(cor) %>% 
          mean()

# how often is a method top when subsampling
top_method_rank_match <- read_rds(result_file("task_sample_ranks_match.rds", "11-evaluation_robustness")) %>%
filter(method_short_name == "slngsht") %>% 
filter(n_tasks == round(max(n_tasks)/10)) %>%
pull(rank_match)

# how many datasets are necessary to get a method on top
random_method_max_tasks_top <- read_rds(result_file("task_sample_max_tasks_top.rds", "11-evaluation_robustness")) %>% 
filter(method_short_name == "wishbone") %>% 
pull(max_n_tasks_top)
```

We further investigated the robustness of our evaluation across datasets and dataset sources. Method performance was moderately to highly correlated (0.5-0.9 depending on the trajectory type) between real and synthetic datasets (`r ref("sfig", "robustness_comparison", "a")`), confirming the relevance of synthetic data, and the accuracy of the real datasets' gold standard. Furthermore, while performance was strongly dependent on individual datasets, the ordering of methods was still robust, with a majority of the methods significantly outperforming the next method in the ranking (`r ref("sfig", "robustness_comparison", "b")`). To further illustrate this, we looked at the scores in a randomly selected 10% subset of all datasets. The overall ordering was still highly correlated with the ordering on all datasets (average spearman's correlation of  `r round(correlation_tenth, 2)`), and the top method, Slingshot, was the top method in `r scales::percent(top_method_rank_match)` of random dataset subsets. On the other hand, for nearly every method it was possible to find a subset of datasets on which it on average performed the best  (`r ref("sfig", "robustness_comparison", "c")`). For example, for Wishbone, a method ranked in the middle, we found a subset of `r random_method_max_tasks_top` datasets on which it would be the top ranking method. Altogether, this demonstrates that almost every method can be the best on a particular set of datasets, and that a large number of datasets, as we included in our evaluation, is necessary to robustly compare the overall performance of methods.

```{r}
add_sfig(
  figure_file(
    "robustness.svg",
    "11-evaluation_robustness"
  ),
  "robustness_comparison", 
  "The robustness of our evaluation.",
  "a) Comparison of method performance between real and synthetic datasets across trajectory types. Given in the top-left corner of each panel is the Pearson's correlation. \"All\" includes the performance on all individual trajectory types, while \"Overall\" includes the performance averaged over all trajectory types. b) Violin plots of the overall performance on each individual dataset. For each method, the arch indicates the first method with which a method performs significantly higher (using a pairwise wilcoxon test corrected for multiple testing; p-value < 0.05). c) Estimates of the size of largest subset of datasets on which the method would be the average top performer. For each method, starting from all datasets, we removed datasets one by one by order of their performance, and tested whether this method ranks first when aggregating the scores over these datasets. This number is therefore only a lower bound, and the actual size of largest subset could be higher."
)
```

@@

<!-- Inverse relation between edge flip and correlation -->
In most cases, the methods which were specifically designed to handle a particular trajectory type indeed performed better on data containing this particular trajectory type (`r ref("fig", "results_overview", "b")`). These methods typically predicted a better topology and cellular neighbourhood, compared to methods which were not able to handle the particular trajectory type. However, the accuracy of cell ordering typically followed the opposite pattern, where methods restricted to linear trajectory types, such as Embeddr, SCORPIUS and Topslam, produced the best ordering, irrespective of whether the dataset contained a linear trajectory or more complex trajectories. This indicates that current non-linear methods potentially contain less accurate ordering algorithms. A combination of the ordering algorithms from the top linear methods, together with the topology inference from top non-linear methods, could therefore be a possibility for future research.

<!-- Topological complexity versus performance -->
```{r}

# sensitivity of topology prediction
topology_sensitivity <- read_rds(result_file("topology_sensitivity.rds", "8-compare_topology"))
```

Despite their similar overall performance, the topologies predicted by the top methods differed considerably. Trajectories detected by Slingshot and cellTree tended to contain less branches, while those detected by TSCAN and AGA Pseudotime gravitated towards more complex topologies (`r ref("fig", "topology_complexity_comparison")`). AGA Pseudotime for example frequently predicted a tree topology, even when a non-branching topology was present in the data. Indeed, when we assessed how often a method inferred the correct topology, Slingshot and TSCAN were good at correctly predicting linear, bifurcating and converging trajectories, while Monocle DDRTree most frequently predicted the correct tree topology (`r ref("stable", "topology_sensitivity")`). Although the top methods in the ranking were capable at discriminating linear from bifurcating trajectories (with TSCAN predicting the correct topology in respectively `r scales::percent(filter(topology_sensitivity, method_id == 'tscan', trajectory_type == 'directed_linear')$perc_perfect)` and `r scales::percent(filter(topology_sensitivity, method_id == 'tscan', trajectory_type == 'bifurcation')$perc_perfect)` of datasets), the accuracy when more complex tree trajectories were present in the data was still very low, with the best methods, Monocle DDRTree and SCUBA, predicting the correct tree topology in respectively `r scales::percent(filter(topology_sensitivity, method_id == 'mnclddr', trajectory_type == 'rooted_tree')$perc_perfect)` and `r scales::percent(filter(topology_sensitivity, method_id == 'scuba', trajectory_type == 'rooted_tree')$perc_perfect)` of the cases  (`r ref("stable", "topology_sensitivity")`). Inferring the correct topology without parameter tuning is therefore still an open challenge.

```{r echo=FALSE, results="asis"}
add_fig(
  fig_path = figure_file("topology_complexity_comparison.svg", experiment_id = "8-compare_topology"),
  ref_id = "topology_complexity_comparison", 
  caption_main = "Comparison of the size (number of nodes and number of edges) of the topology predicted by the top TI methods, compared to the size of the gold standard.", 
  caption_text = "Shown are the distributions of the difference in size between predicted and gold standard topologies. A positive difference means that the topology predicted by the method is more complex than the one in the gold standard."
)
```

```{r, results='asis' }
add_stable(
  read_rds(figure_file("topology_sensitivity_table.rds", "8-compare_topology"))[[params$table_format]],
"topology_sensitivity",
"Sensitivity of topology prediction of each TI method. Given is how often a method is able to correctly predict the topology of the trajectory present in the dataset, within a particular trajectory type."
)
```

<!-- Additional factors which could influence performance -->
Apart from the trajectory types, we also investigated several other factors that could influence the performance of a method. Although we provided prior information to those methods that require it, we did not observe any major difference in method performance between these methods and those that did not receive any prior information (`r ref("fig", "results_overview", "a")`). When we looked at the algorithmic steps shared by different methods, we found that those which included principal curves (such as Slingshot and SCORPIUS), k-means (which include Slingshot and several other top scoring methods) and some graph building (which include almost every top scoring method) tended to have a slightly higher performance, while those which included tSNE had a lower performance ( `r ref("stable", "component_importance")`), although these tendencies were not significant.

```{r}
add_stable(
  read_rds(figure_file(
    "component_importance.rds",
    "6-performance_predictors"
  ))[[params$table_format]], 
  "component_importance", 
  "Importance values for algorithmic components. The overall performance was compared between methods containing a particular algorithmic component, and we calculated both a p-value using a two-tailed Mann–Whitney U test (corrected for multiple testing using the method of Benjamini-Hochberg) and the increase in node purity importance measure using random forest classification."
)
```

@@

While not directly related to the accuracy of the inferred trajectory, the quality of the implementation is also an important evaluation metric [@_doesyourcode_2018]. To assess this, we scored each method using a transparent checklist of important scientific and software development practices, including software packaging, documentation, automated code testing, and publication into a peer-reviewed journal (`r ref("stable", "qc_checks")`). This allowed us to rank each method based on its user friendliness, developer friendliness, and potential broad applicability on new unseen datasets.

We found that most methods fulfilled the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "results_overview", "c")` and `r ref("fig", "implementation_qc_overview")`). While recent methods had a slightly better quality score than older methods, several quality aspects were consistently lacking for the majority of the methods (`r ref("fig", "implementation_qc_overview", " right")`) and we believe that these should receive extra attention from developers. Although these outstanding issues cover all five categories, code assurance and documentation in particular are problematic areas, notwithstanding several studies pinpointing these as good practices [@wilson_best_2014; @artaza_top10metrics_2016].

```{r echo=FALSE, results="asis"}
add_fig(
  fig_path = figure_file("implementation_qc_overview.svg", experiment_id = "4-method_characterisation"),
  ref_id = "implementation_qc_overview", 
  caption_main = "Overview of the quality control results for every method.", 
  caption_text = glue::glue("Shown is the score given for each method on every item from our quality control score sheet ({ref('stable', 'qc_checks')}). Each aspect of the quality control was part of a category, and each category was weighted so that it contributed equally to the final quality score. Within each category, each aspect also received a weight depending on how often it was mentioned in a set of papers discussing good practices in tool development and evaluation. This is represented in the plot as the height on the y-axis. Top: Average QC score for each method. Right: The average score of each quality control item. Shown into more detail are those items which had a score lower than 0.5.")
)
```

We observed no clear relation between method quality and method performance (`r ref("fig", "results_overview")`). We could also not find any quality aspect which was significantly associated with method performance.

@@

```{r echo=FALSE, results="asis"}
add_fig(
  fig_path = figure_file("tree.svg", experiment_id = "7-user_guidelines"),
  ref_id = "user_guidelines", 
  caption_main = "Practical guidelines for method users.", 
  caption_text = "As the performance of a method most heavily depends on the topology of the trajectory, the choice of TI method will be primarily influenced by the user's existing knowledge about the expected topology in the data. We therefore devised a set of practical guidelines, which combines the method's performance, user friendliness and the number of assumptions a user is willing to make about the topology of the trajectory. Methods to the right are ranked according to their performance on a particular (set of) trajectory type. Further to the right are shown the user friendliness scores (++: ≥ 0.95, +: ≥ 0.85, ± ≥ 0.75, - ≥ 0.65), overall performance (++: top method, +: difference between top method's performance ≥ -0.05, ±: ≥-0.2, -: ≥ -0.5) and required prior information."
)
```

@@

```{r echo=FALSE, results="asis"}
add_fig(
  fig_path = figure_file("bifurcating_example.svg", experiment_id = "14-example_predictions"),
  ref_id = "example_predictions", 
  caption_main = "Demonstration of how a common framework for TI methods facilitates broad applicability using an example dataset.", 
  caption_text = "a) The dataset consists of fibroblasts being reprogrammed towards neurons and myocytes, and thus contains a bifurcating dynamic process  [@treutlein_dissectingdirectreprogramming_2016]. A multi-dimensional scaling of dataset is shown at the bottom. MEF: mouse embryonic fibroblast b) A common visualisation of the predicted trajectories allows to easily compare the similarities and differences between predictions. The predictions are in line with earlier results in that Slingshot underestimates the complexity of the topology, AGA Pseudotime and Monocle DDRTree overestimate the complexity, and TSCAN is able to correctly identify the trajectory topology. c) More insight can be gained into the ordering of the cells by placing the branches one after the other. The predicted topology is shown in the bottom of each plot. d) Method-specific visualisations provide insight into the inner workings of a method's methodology and into how the trajectory was constructed. e) Our metrics can be used to compare the output to a gold standard or to each other. In this case, while only TSCAN and cellTree correctly predict the topology, the ordering of Slingshot and AGA Pseudotime on this dataset corresponds better to the expected ordering."
)
```

@@

# Online Methods

## Trajectory inference methods
<!-- Overview -->
We gathered a list of `r sum(implementations$contains_ti, na.rm=TRUE)` trajectory inference methods (`r ref("stable", "implementations")`), by searching the literature for "trajectory inference" and "pseudotemporal ordering", and based on two existing lists found online [@davis_awesomesinglecelllistsoftware_2018; @gitter_singlecellpseudotimeoverviewalgorithms_2018].
A continuously updated list can also be found [online](https://goo.gl/FnsTxG)). We welcome any contributions by creating an issue at [github.com/dynverse/dynverse/issues](https://github.com/dynverse/dynverse/issues).

>>>>

---
title: "A comparison of single-cell trajectory inference methods: towards more accurate and robust tools"
output: dynbenchmark::pdf_manuscript
---

```{r setup, include=FALSE}
library(dynbenchmark)
library(tidyverse)

tools <- read_rds(result_file("tools.rds", "03-methods"))
methods <- read_rds(result_file("methods.rds", "03-methods"))
```

@@

```{r}
n_tools <- tools %>% filter(type == "algorithm") %>% nrow()
n_methods_evaluated <- nrow(filter(methods, evaluated, type %in% c("algorithm", "algorithm_test")))
n_control_methods_evaluated <- nrow(filter(methods, evaluated, type %in% c("control", "control_test")))
n_datasets_real <- 10
n_datasets_synthetic <- 20
```
# Abstract

Recent technological advances allow unbiased investigation of cellular dynamic processes, by generating -omics profiles of thousands of single cells and computationally ordering these cells along a trajectory. Since 2014, at least → `r n_tools` ←tools for trajectory inference have been developed, but due to high variability in their inputs and outputs, these methods are difficult to compare. As a result, a comprehensive assessment of the performance of trajectory inference methods is still lacking, and more importantly also guidelines for new users to the field.

In this study, we evaluated a total of → `r n_methods_evaluated` ←methods in terms of cellular →ordering, topology, scalability, and user friendliness←. Our results indicate that existing methods are typically better at ordering the cells rather than inferring the correct trajectory topology, and that there is no "one-fits-all" method. Instead, the choice of method depends on the trajectory topology present in the data. →To aid users with selecting the most appropriate method, we therefore developed a set of guidelines, available as an interactive app at `r print_url("http://35.239.15.98")`,← While the field certainly has matured significantly since its inception, further progress is still necessary in order to cope with increasingly complex and novel use cases. We hope to spearhead such developments and the field as a whole, by making our evaluation pipeline easily extensible and freely available at `r print_url("https://www.github.com/dynverse/dynverse")`.

@@

<!-- §2: Plethora of new TI methods -->
A plethora of TI methods has been developed over the last years, and even more are being created every month (`r ref("stable", "tools")`). →In several repositories listing single-cell tools,  omictools.org [@henry_omictoolsinformativedirectory_2014], the "awesome single cell software" list [@davis_awesomesinglecelllistsoftware_2018] and scRNA-tools.org [@zappia_exploringsinglecellrnaseq_2017], TI methods are one the largest categories.← All methods have their own unique set of characteristics in terms of underlying algorithm, the prior information they require and the outputs they produce. Two of the most distinctive differences between TI methods are whether they fix the topology of the trajectory or not, and what type(s) of graph topologies they can detect. Early TI methods typically fixed the topology algorithmically (e.g. linear [@bendall_singlecelltrajectorydetection_2014; @schlitzer_identificationcdc1cdc2committed_2015; @shin_singlecellrnaseqwaterfall_2015; @campbell_bayesiangaussianprocess_2015] or bifurcating [@haghverdi_diffusionpseudotimerobustly_2016; @setty_wishboneidentifiesbifurcating_2016]), or through parameters provided by the user [@trapnell_dynamics_2014; @matsumoto_scoupprobabilisticmodel_2016]. These methods therefore mainly focused on correctly ordering the cells along the fixed topology. →Most recent methods also infer the topology computationally, which increases the difficulty of the problem at hand, but allows an unbiased identification of both the ordering inside a branch and the topology connecting these branches.←

```{r, eval = FALSE}
add_stable(
  read_rds(result_file("tools_table.rds", "03-methods"))[[params$table_format]],
"tools",
paste0("Overview of the trajectory inference methods included in this study. ", glue::glue_data(non_inclusion_reasons, "{footnote}: {long}") %>% glue::glue_collapse("; "))
)
```

@@

<!-- §4: Our study -->
In this study, we performed a comprehensive evaluation of  →`r n_methods_evaluated`← TI methods (`r ref("fig", "ti_evaluation_overview", "a")`, `r ref("sfig", "overview_evaluation_expanded")`). The inclusion criterion for TI methods was primarily based on their free availability and presence of a programming interface (`r ref("stable", "tools")`). →The evaluated each methods on four core aspects: (i) their accuracy given a gold or silver standard on `r n_datasets_real` real and `r n_datasets_synthetic` synthetic datasets, (ii) the user friendliness of the provided software and documentation, (iii) their scalability with both the number of cells and features, and (iv) their stability across reruns stability. Finally, we combined these metrics to provide a set of practical guidelines for users. ←

```{r}
add_fig(
  fig_path = "figures/overview_evaluation_v4.svg",
  ref_id = "ti_evaluation_overview", 
  caption_main = "Overview of several key aspects of the evaluation.", 
  caption_text = "a) A schematic overview of our evaluation pipeline. b) In order to make the trajectories comparable to each other, a common trajectory model was used to represent gold standard trajectories from the real and synthetic datasets, as well as any predicted trajectories from TI methods. c) Trajectories are automatically classified into one of nine trajectory types, with increasing complexity. d) Three metrics were defined in order to assess the similarity in cell ordering, cell neighbourhood, and topology, for any two trajectories.",
   width = 12,
  height = 8, 
)
```

```{r}
add_sfig(
  fig_path = "figures/overview_evaluation_expanded_v1.svg",
  ref_id = "overview_evaluation_expanded",
  caption_main = "Extended overview of our evaluation pipeline.", 
  caption_text = "From literature we extracted a quality control checklist, a list of trajectory inference methods, a set of real datasets containing a trajectory and real regulatory networks used to generate the synthetic data. We created a wrapper of each method, so that its output was transformed into a common probabilistic trajectory model, and used these to infer trajectories on all real and synthetic datasets using their default parameters. Using several similarity metrics, we compared the gold standard of the real and synthetic datasets with the inferred trajectories. We also evaluated the quality of each method using the quality control checklist, and used both the benchmark results and quality control results to produce a final set of guidelines for method's users.", 
  width = 15, 
  height = 7.5
)
```

@@

<!-- Evaluation overview -->
In order to make gold standard trajectories and predicted trajectories directly comparable to each other, we developed a common probabilistic model for representing trajectories from all possible sources (`r ref("fig", "ti_evaluation_overview", "b")`). In this model, the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of connected milestones. The milestone network is automatically classified into one of the nine predefined trajectory types (`r ref("fig", "ti_evaluation_overview", "c")`), which go from the very basic topologies (linear, cyclical and bifurcating) to the most complex (cyclic graphs and disconnected graphs). →We defined several possible metrics to compare a prediction to the reference trajectory, and based on an analysis of their robustness and conformity to a set of rules (`r ref("snote", "metrics")`), we chose a set of 4 metrics each assessing a different aspect of a trajectory: the cell ordering (`r label_metric("correlation")`), the topology (`r label_metric("him")`), the quality of differentially expressed features  (`r label_metric("featureimp_cor")`), and the quality of clustering of the cells into branches  (`r label_metric("F1_branches")`) (`r ref("fig", "ti_evaluation_overview", "d")`).←The data compendium consisted of both synthetic datasets, which offer the most exact gold standard, and real datasets, which offer the highest biological relevance. These real datasets come from a variety of single-cell technologies, organisms, and dynamic processes, and contain several types of trajectory topologies (`r ref("stable", "datasets")`). →For synthetic datasets we used several data simulators, including our own which simulates gene regulatory networks using a thermodynamic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011], and subsequently simulates single-cell profiling experiments by matching the distributions of the synthetic data with real reference datasets.←

<!-- Overall evaluation results -->
```{r echo=FALSE, results="asis"}
add_fig(
  fig_path = result_file("overview.pdf", experiment_id = "08-summary"),
  ref_id = "results_overview", 
  caption_main = glue::glue("Overview of the results on the evaluation of {n_methods_evaluated} TI methods and {n_control_methods_evaluated} control methods."), 
  caption_text = "a) The methods were characterised according to the most complex trajectory type they can infer, whether or not the inferred topology is constrained by the algorithm or a parameter, and which prior information a method requires or can optionally use. b) The methods are ordered according to the overall score achieved both in the benchmark and QC evaluation. `r {n_control_methods_evaluated}` control methods were also included as an indication of baseline performance c) Also shown are the aggregated scores per metric, source and trajectory type, as well as the average execution time across all datasets and the percentage of executions in which no output was produced. d) Performance in the quality control evaluation is highly variable, even amongst the highest ranked methods according to the benchmark evaluation. Also listed are the quality control scores aggregated according to practicality and the different categories.",
  width = 15,
  height = 20
)
```

<!-- Evaluation robustness -->

@@

<!-- Inverse relation between edge flip and correlation -->

<!-- Topological complexity versus performance -->

## Method stability and scalability

@@

@@

```{r echo=FALSE, results="asis"}
add_fig(
  fig_path = ggplot(),#figure_file("tree.svg", experiment_id = "7-user_guidelines"),
  ref_id = "user_guidelines", 
  caption_main = "Practical guidelines for method users.", 
  caption_text = "As the performance of a method most heavily depends on the topology of the trajectory, the choice of TI method will be primarily influenced by the user's existing knowledge about the expected topology in the data. We therefore devised a set of practical guidelines, which combines the method's performance, user friendliness and the number of assumptions a user is willing to make about the topology of the trajectory. Methods to the right are ranked according to their performance on a particular (set of) trajectory type. Further to the right are shown the user friendliness scores (++: ≥ 0.95, +: ≥ 0.85, ± ≥ 0.75, - ≥ 0.65), overall performance (++: top method, +: difference between top method's performance ≥ -0.05, ±: ≥-0.2, -: ≥ -0.5) and required prior information."
)
```

@@

```{r echo=FALSE, results="asis"}
add_fig(
  fig_path = ggplot(),#figure_file("bifurcating_example.svg", experiment_id = "14-example_predictions"),
  ref_id = "example_predictions", 
  caption_main = "Demonstration of how a common framework for TI methods facilitates broad applicability using an example dataset.", 
  caption_text = "a) The dataset consists of fibroblasts being reprogrammed towards neurons and myocytes, and thus contains a bifurcating dynamic process. A multi-dimensional scaling of dataset is shown at the bottom. MEF: mouse embryonic fibroblast b) A common visualisation of the predicted trajectories allows to easily compare the similarities and differences between predictions. The predictions are in line with earlier results in that Slingshot underestimates the complexity of the topology, AGA Pseudotime and Monocle DDRTree overestimate the complexity, and TSCAN is able to correctly identify the trajectory topology. c) More insight can be gained into the ordering of the cells by placing the branches one after the other. The predicted topology is shown in the bottom of each plot. d) Method-specific visualisations provide insight into the inner workings of a method's methodology and into how the trajectory was constructed. e) Our metrics can be used to compare the output to a gold standard or to each other. In this case, while only TSCAN and cellTree correctly predict the topology, the ordering of Slingshot and AGA Pseudotime on this dataset corresponds better to the expected ordering."
)
```

@@

# Online Methods
## Data and code availability
→ 
The real and synthetic datasets used in this study are deposited on Zenodo ([doi.org/10.5281/zenodo.1211533](https://doi.org/10.5281/zenodo.1211533)) [@cannoodt_goldstandardsinglecell_].

The code to fully reproduce the results from this study has been split into several R packages and is available on github at `r print_url("github.com/dynverse/dynverse")`. These include packages for wrapping trajectories (`r print_url("github.com/dynverse/dynwrap")`), all trajectory inference methods (`r print_url("github.com/dynverse/dynverse")`), the evaluation metrics (`r print_url("github.com/dynverse/dyneval")`) and the guidelines app packages (`r print_url("github.com/dynverse/dynguidelines")`). The main analysis repository is available at `r print_url("github.com/dynverse/dynbenchmark")` and is divided into several experiments. In this repository, we provide READMEs for both the scripts and the results, which can be easily browsed and explored on the github website. 
←
## Trajectory inference methods
<!-- Overview -->
We gathered a list of `r n_tools` trajectory inference tools (`r ref("stable", "tools")`), by searching the literature for "trajectory inference" and "pseudotemporal ordering", and based on two existing lists found online [@davis_awesomesinglecelllistsoftware_2018; @gitter_singlecellpseudotimeoverviewalgorithms_2018].
We welcome any contributions by creating an issue at →`r print_url("github.com/dynverse/dynmethods/issues")`←.

<!-- Inclusion criterion -->
Methods were excluded from the evaluation based on several criteria: `r glue::glue_collapse(glue::glue("({non_inclusion_reasons$footnote}) {non_inclusion_reasons$long}"), ", ")`. This resulted in the inclusion of  `r n_methods_evaluated` methods in the evaluation.
